---
title:  "CGIAR co-Authorhips"
author: "IFPRI/CIAT"
date:   "December, 2015. Last updated on `r Sys.Date()`. DRAFT, do not use or cite."
output:
  html_document:
    fig_height: 5
    toc: true
    toc_float: true    
    fig_caption: true  
---

Aim is to create a dynamic network visualization of CGIAR authors and
co-authors showing citation details and sub-communities (key actors and
links) with relationship strenghts representing number of co-authored papers.
GoogleScholar (GS) does not yet provide an API yet (not an official Google
product).

Found 2 general-purpose scraping libraries `rvest` and `scrapeR`.
`scrapeR` uses straight xpath, rvest supports `%>%` piping. Inspired from blog
post at http://datascienceplus.com/google-scholar-scraping-with-rvest/.
`networkD3` and `igraph` may be used to generate dynamic network graphs with
tutorial at http://kateto.net/network-visualization.

`igraph` needs 2 data.frames:

* *links* - describes the edges of the network. Its first two columns are the
  IDs of the source and the target node for each edge. The following columns are
  edge attributes (weight, type, label, or anything else).
* *nodes* - starts with a column of node IDs. Any following columns are
  interpreted as node attributes.


## Scrape CGIAR-affiliated Authors

We'll use Google Scholar citations as a starter. GS does not make it easy to
scrape the entire catalog for specific keywords (cannot page through query
results using URL parameters), so we need an iterative approach instead. Also
note that a majority of authors are not "verified" (records only show
unstructured list of authors in `Lastname, F` format not hyperlinked to their
profile page or not showing any e-mail domain name). Few author profiles include
list of co-authors, but only when entered manually. This makes GS difficult to
scrape altogether.

Using the list of co-authors on author's profile pages would bias the graph
towards authors with complete profiles and not necessarily a higher number of
publications or citations.


```{r setup, results='hide'} 
 
library(data.table)
library(rvest)
library(stringr)
library(igraph)
library(networkD3)

setwd("~/Projects/hc-shiny/cgiarPubs")

# Load data snapshot
load("./tmp/GoogleScholar.RData")

```


```{r, eval=F}

# Get GS IDs and attributes of CGIAR authors over 2000-2015 period
gs.url <- "https://scholar.google.com/citations?view_op=view_org&hl=en&org=5225706230162989193&before_author="
gs.url.author <- "https://scholar.google.com/citations?hl=en&user="
gs.url.colleagues <- "https://scholar.google.com/citations?view_op=list_colleagues&hl=en&user="
gs.root.author <- "WiY9Ku0AAAAJ" # top cited CGIAR author id
page <- read_html(paste0(gs.url, gs.root.author))

# Helper functions
scrape.org <- function(id) {
  page <- read_html(paste0(gs.url, id))
  cg.author <- page %>% html_nodes(css=".gsc_1usr_name") %>% html_text()
  cg.id <-  page %>% html_nodes(xpath="//h3[@class='gsc_1usr_name']/a/@href") %>% html_text()
  cg.id <- str_replace(str_replace(cg.id, fixed("/citations?user="), ""), "&hl=en&oe=ASCII", "")
  cg.cited <- page %>% html_nodes(css=".gsc_1usr_cby") %>% html_text()
  cg.cited <- as.integer(str_replace(cg.cited, "Cited by ", ""))
  cg.title <- page %>% html_nodes(css=".gsc_1usr_aff") %>% html_text()
  cg.key <- page %>% html_nodes(css=".gsc_1usr_int")
  cg.key <- lapply(cg.key, html_nodes, css="a")
  cg.key <- lapply(cg.key, html_text)
  tmp <- data.table(id=cg.id, fullName=cg.author, cited=cg.cited, title=cg.title, keywords=cg.key)
  return(tmp)
}

scrape.author <- function(id) {
  page <- read_html(paste0(gs.url.author, id))
  cg.author <- page %>% html_node(css="#gsc_prf_in") %>% html_text()
  cg.cited <- page %>% html_node(css=".gsc_rsb_std") %>% html_text()
  cg.cited <- as.integer(cg.cited)
  cg.title <- page %>% html_node(css=".gsc_prf_il") %>% html_text()
  cg.key <- page %>% html_nodes(css=".gsc_prf_ila")
  cg.key <- sapply(cg.key, html_text)  
  cg.cgiar <- page %>% html_node(css="#gsc_prf_ivh") %>% html_text()
  cg.cgiar <- str_detect(cg.cgiar, "cgiar.org")
  tmp <- data.table(id=id, fullName=cg.author, cited=cg.cited, title=cg.title, cgiar=cg.cgiar)
  tmp[, keywords := paste(tolower(cg.key), collapse="; ")]
  return(tmp)
}

scrape.coauthors <- function(id) {
  page <- read_html(paste0(gs.url.colleagues, id))
  cg.author <- page %>% html_nodes(css=".gsc_1usr_name") %>% html_text()
  cg.id <-  page %>% html_nodes(xpath="//h3[@class='gsc_1usr_name']/a/@href") %>% html_text()
  cg.id <- str_replace(str_replace(cg.id, fixed("/citations?user="), ""), "&hl=en&oe=ASCII", "")
  if (length(cg.id)>0) tmp <- data.table(id=id, ids=cg.id, coFullName=cg.author)
  else tmp <- data.table()
  return(tmp)
}

```

```{r, eval=F}

# Start scraping from root author down through his CGIAR co-authors
top.id <- scrape.coauthors(gs.root.author)$ids

# Init
cg.author <- data.table(id=NA, fullName=NA, cited=NA, title=NA, cgiar=NA, keywords=NA)
cg.coauthor <- data.table(id=NA, ids=NA)

# Start from root author and continue to scrape through list of co-authors
for (i in 1:500) {
  j <- 1
  while (top.id[j] %in% cg.author$id) {
    # Move down each list of co-authors
    if (j < length(top.id))  j <- j+1 else break
  }
  if (top.id[j] %in% cg.author$id) {
    # Don't re-scrape the same author
    top.id <<- cg.coauthor[!ids %in% cg.author$id, ids]
    j <- 1
  }
  cat(i, ":", top.id[j], "\n")
  tmp <- scrape.author(top.id[j])
  tmp1 <- scrape.coauthors(top.id[j])
  cg.coauthor <- rbind(cg.coauthor, tmp1, fill=T)
  setkey(cg.coauthor, id, ids)
  cg.coauthor <- unique(cg.coauthor)  
  tmp[, coauthors := nrow(tmp1)]
  cg.author <- rbind(cg.author, tmp, fill=T)
  setkey(cg.author, id)
  cg.author <- unique(cg.author)
  if (nrow(tmp1)>0) {
    # Rank co-authors
    tmp2 <- lapply(tmp1$ids, scrape.author)
    tmp2 <- rbindlist(tmp2)
    tmp2 <- tmp2[order(-cited)]
    # Scrape CGIAR-affiliated co-authors first
    if (length(top.id)>0) top.id <<- tmp2[cgiar==T, id]
    #if (length(top.id)==0) top.id <<- tmp2$id
  }
}

# Remove duplicate authors
setkey(cg.author, id)
cg.author <- unique(cg.author)
cg.author <- cg.author[!is.na(id)]

# Clean up list of keywords
cg.keywords <- paste(cg.author$keywords, collapse="; ")
cg.keywords <- unlist(str_split(cg.keywords, "; "))
cg.keywords <- unique(cg.keywords)
cg.keywords <- cg.keywords[!cg.keywords %in% c("", " ", NA)]

# Show first 10 extracted authors
head(cg.author, 10)

```

Using 500 page scrapes the steps above extracted `r nrow(cg.coauthor)`
co-author relationships, out of which we identifed `r nrow(cg.author)` unique
authors.

## Graph CGIAR co-Authorships

First let's graph the relationships between authors and co-authors, trying to
identify CGIAR-affiliated and non-CGIAR authors.

```{r, eval=F, results='hide'}

## Prepare data for igraph
nodes <- cg.author
nodes[, ID := 0:(.N-1)]
nodes[, size := scale(cited, center=F)*4]
nodes[, group := factor(cgiar, labels=c("Partners", "CGIAR"))]
links <- cg.coauthor[id %in% nodes$id]
links <- links[ids %in% nodes$id]
setkey(links, id)
links$from <- nodes[links][, ID]
setkey(links, ids)
links$to <- nodes[links][, ID]
links[, value := 3]
setcolorder(nodes, c(8:10, 1:7))
setcolorder(links, c(4,5,6,1:3))

```

```{r}

# Network plots
net <- graph.data.frame(links, nodes, directed=F)
plot(net, 
  edge.arrow.size=.2, 
  edge.color="orange",
  vertex.color=ifelse(V(net)$cgiar, "green", "orange"), 
  vertex.label.color="black",
  vertex.frame.color="white",
  vertex.size=log10(V(net)$cited),
  vertex.label=V(net)$fullName,
  vertex.label.cex=.5)

```

Same as above with D3 interactive network graph. We'd like to color-code CGIAR
center affiliations but this requires cleaning up organization records.

```{r, results='hide'}

# Add node attributes (incl. cluster membership via random walk)
tmp <- membership(cluster_walktrap(net))
nodes[, label := paste(fullName, title, keywords, sep=" - ")]
setkey(nodes, ID)
nodes[J(as.integer(names(tmp))), cluster := as.vector(tmp)]

```

```{r}

# Color-code affiliation
forceNetwork(links, nodes,
  Source="from", Target="to",
  NodeID="label", Nodesize="size",
  linkColour="#666", 
  fontSize=12, fontFamily="arial",
  Group="group", 
  colourScale=JS("d3.scale.category10()"),
  zoom=T, legend=T)

```

 


```{r}

# Color-code cluster
forceNetwork(links, nodes,
  Source="from", Target="to",
  NodeID="label", Nodesize="size",
  linkColour="#666", 
  fontSize=12, fontFamily="arial",
  Group="cluster", 
  colourScale=JS("d3.scale.category20()"),
  zoom=T, legend=T)

```

 

**TODO** We would also need to consolidate the list of organizations, identify
country of affiliation, consolidate keywords using the latest CGIAR
ontologies, then identify topical clusters of authors. Also need to look through
other repositories (e.g. [Web of
Science](http://ipscience.thomsonreuters.com/product/web-of-science/)).


```{r, eval=F}

# Save snapshot
rm(list=ls()[ls() %like% "tmp"])
save.image("./tmp/GoogleScholar.RData")

```

