---
title:  "Scrape CGIAR Publications from ScienceDirect, 1980-2016"
author: "IFPRI/CIAT"
date:   "October 2016. Last updated on `r Sys.Date()`. DRAFT, do not use or cite."
output:
  html_document:
    fig_height: 8
    fig_width: 8
    toc: true
    toc_float: true    
    fig_caption: true  
---

Extract basic metadata records from CGIAR centers from ScienceDirect
on-line catalog. Identify centers, authors and co-author affiliations
when feasible. Explore trends in authorships especially changes between pre- and
post-CGIAR Reform (around 2010).

We queried ScienceDirect journals and books for author affiliations matching
CGIAR centers (incl. retired centers, ILRAD, ISNAR, ILCA, and INIBAP). Note that
we could also hit **ScienceDirect API** at
http://api.elsevier.com/documentation/SCIDIRSearchAPI.wadl#N1002B or
**Scopus API** at https://dev.elsevier.com/ but these APIs don't seem to return
author ids (only primary author and unstructured list of co-authors).

Note that R package `RefManageR` may also be used to query CrossRef API,
GoogleScholar, and Zotero -- could be useful as well. Package `rscopus` provides
a few helper methods to interact with Elsevier Scopus APIs.


```{r}

library(data.table)
library(RefManageR)
library(rvest)
library(stringr)
library(ggplot2)
library(pander)
library(rscopus)

setwd("~/Projects/hc-shiny/cgiarPubs")

# Load data snapshot
load("./tmp/ScienceDirect.RData")

```

# Scrape CGIAR publications from ScienceDirect

```{r, eval=F}

# Load Bibtex records downloaded from ScienceDirect, one file per center
f <- list.files("./in/ScienceDirect", "*.bib", full.names=T)
bib <- lapply(f, ReadBib, check=F)
names(bib) <- str_replace(str_replace(basename(f), "science_", ""), ".bib", "")

```

```{r, warning=F}

# Show first 2 records
pander(bib[[1]][1:2])

```

Note that author e-mail and affiliation are typically missing.

```{r, eval=F}

# Add `institution` field from file name
cg <- c("Africa Rice", 
  "Bioversity International", "Bioversity International", 
  "CIAT", "CIFOR", "CIMMYT", "CIP", "ICARDA",        
  "ICRAF", "ICRISAT", "IFPRI", "IFPRI",     
  "IITA", "ILRI", "ILRI", "IRRI", "IWMI", "WorldFish")

# Flatten all records
refs <- lapply(bib, as.data.frame)
refs <- lapply(refs, data.table)
for (i in 1:length(cg)) refs[[i]][, institution := cg[i]]
refs <- rbindlist(refs, fill=T)

# Remove duplicates
setkey(refs, url)
refs <- unique(refs)

# Make arbitrary unique publication IDs
refs[, 
  uid := tolower(paste(abbreviate(institution, minlength=3), str_pad(.I, 4, pad="0"), sep="-")),
  by=institution]

```

```{r}

# Print records
pander(refs[[1]][1:10])

# Tally by CGIAR center
ggplot(refs[, .N, keyby=institution], 
  aes(institution, N)) +
  geom_bar(stat="identity", color="#ffffff", alpha=0.6) +
  labs(x="", y="", title="CGIAR Publications by center (Elsevier)") +
  coord_flip()

# Tally by publication year
ggplot(refs[, .N, keyby=.(year=as.Date(year, "%Y"), Center=institution)],
  aes(year, N, fill=Center)) +
  geom_bar(stat="identity", alpha=0.6) +
  labs(x="", y="", title="CGIAR Publications by year (Elsevier)")

# Tally by journal (note these are Elsevier journals only)
pander(refs[!is.na(journal), .N, keyby=journal])

```

In total `r nrow(refs)` CGIAR-affiliated publications (incl. journal
articles and manuscripts) are extracted from ScienceDirect.
`r refs[!is.na(doi)]` have a unique DOI number, that may be used to query web
services for additional details, incl. full-text articles.

We further need to structure the list of authors, co-authors and their
affiliation, as this information is not available from the Bibtex dumps. We use
`rvest` to further scrape this information from ScienceDirect.

# Scrape CGIAR authors and publications from Scopus APIs

Instead of importing records from ScienceDirect Bibtex dumps we could also try
to use John Muschelli's R package `rscopus` to query Elsevier Scopus APIs
documented at https://cran.r-project.org/web/packages/rscopus/rscopus.pdf.

```{r}

# Set API key
options(elsevier_api_key=api_key)

# Scopus can be searched against a publication DOI or PII.
# Make PII explicit
refs[, pii := str_replace(url, "http://www.sciencedirect.com/science/article/pii/", "")]

# Use PII query
tmp <- generic_elsevier_api(query="pii(0001706X89900223)", 
  type="search", search_type="scopus", api_key=get_api_key())

# Print response
pander(tmp$content$`search-results`$entry[[1]])

# Use affiliation query
q <- "affil(ifpri)"
tmp <- generic_elsevier_api(query=q, 
  type="search", search_type="affiliation", api_key=get_api_key())
# => not authenticated (not sure that endpoint is working)
tmp <- generic_elsevier_api(query=q, 
  type="search", search_type="scopus", api_key=get_api_key())

# Print first record
pander(tmp$content$`search-results`$entry[[1]])

tmp <- generic_elsevier_api(query="authlast(Rosegrant)", 
  type="author", api_key=get_api_key())
# This author endpoint does not seem to work either: "RESOURCE_NOT_FOUND"

```

Querying against Scopus API works but doesn't return any author UID or OrcID
only full names, and only return the first author affiliation, city and country.
The response does include article citation in field `citedby-count` though,
which may be useful. Also note that Elsevier Author API at endpoint
https://api.elsevier.com/content/search/author does not seem to work right now.

```{r, eval=F}

# Helper - Return article citations, and 1st affiliation city and country from Scopus API
scrapePII <- function(x) {
  q <- paste0("pii(", x, ")")
  tmp <- generic_elsevier_api(query=q, type="search", search_type="scopus", api_key=get_api_key())
  tmp <- data.table(
    pii=x,
    cited=ifelse(is.null(
        tmp$content$`search-results`$entry[[1]]$`citedby-count`),
      as.integer(NA), as.integer(tmp$content$`search-results`$entry[[1]]$`citedby-count`)),
    affname=ifelse(is.null(
        tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affilname`),
      NA, tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affilname`),
    affcountry=ifelse(is.null(
        tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affiliation-country`),
      NA, tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affiliation-country`),
    affcity=ifelse(is.null(
        tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affiliation-city`),
      NA, tmp$content$`search-results`$entry[[1]]$affiliation[[1]]$`affiliation-city`))
  return(tmp)
}

tmp <- scrapePII(refs[1, pii])
for (i in 2:nrow(refs)) tmp <- rbind(tmp, scrapePII(refs[i, pii]))
refs <- cbind(refs, tmp)

```

```{r}

# Summary
pander(summary(refs$cited))

# Show top cited articles
pander(refs[sort(-cited)][1:10, .(title, author, year)])

# Tally citations by center
ggplot(refs[, .(cited=sum(cited, na.rm=T)), keyby=.(Center=institution)],
  aes(Center, cited, fill=Center)) +
  geom_bar(stat="identity", alpha=0.6) +
  labs(x="", y="", title="Citations by center (Elsevier)") +
  coord_flip() + theme(legend.position="none")

```

Instead of using the Scopus API, let's scrape ScienceDirect web pages for
complete author affiliations.


```{r, eval=F}

# Helper - Scrape author affiliation from ScienceDirect by PII
scrapeAuthor <- function(x, y) {
  
  if (is.na(x)) return(data.table(
        uid=NA, 
        name=NA, 
        url=NA, 
        email=NA,
        affid=NA,
        aff=NA, 
        rank=as.integer(NA)))
  
  # There seems to be multiple possible layouts for authors
  p <- read_html(x)
  aut <- p %>% html_nodes(css=".authorGroup li")
  
  if (length(aut) > 0) {
    # Layout #1
    aut <- lapply(aut, function(x) data.table(
          uid=y,
          name=x %>% html_node(css="a.authorName") %>% html_text(),
          url=x %>% html_node(css="a.authorName") %>% html_attr("id"),
          email=x %>% html_node(css=".auth_mail") %>% html_attr("href"),
          affid=x %>% html_node(css="a.auth_aff") %>% html_attr("href")))
    
    aff <- p %>% html_nodes(css=".affiliation li")
    aff <- lapply(aff, function(x) data.table(
          affid=x %>% html_attr("id"),
          aff=x %>% html_node(css="span") %>% html_text()))  
    
  } else {
    # Layout #2
    aut <- p %>% html_nodes(css="span.author")
    aut <- lapply(aut, function(x) data.table(
          uid=y,
          name=x %>% html_node(css=".author-name") %>% html_text(),
          url=x %>% html_node(css=".author-name-link") %>% html_attr("id") ,
          email=x %>% html_node(css=".auth_mail") %>% html_attr("href"),
          affid=x %>% html_node(css=".author-affiliation") %>% html_attr("href")))
    
    aff <- p %>% html_nodes(css=".affiliation")
    aff <- lapply(aff , function(x) data.table(
          affid=x %>% html_attr("id"),
          aff=x %>% html_node(css=".affiliation__text") %>% html_text()))
  }
  
  aut <- rbindlist(aut)
  aff <- rbindlist(aff)
  aut[, affid := str_replace(affid, "#", "")]  
  
  # It can happen that affid is NA, then assign same affiliation to all authors
  aff[is.na(affid), affid := "0"]
  aut[is.na(affid), affid := "0"]
  
  setkey(aut, affid)
  setkey(aff, affid)
  aut[aff, aff := aff]
  
  # Corner case: treat weirdish affiliations starting with "mc" id
  if (aff[1, affid] %like% "mc" | unique(aut$affid)=="0") aut$aff <- aff$aff
  
  # Clean up
  aut[, name := str_replace(name, "            Opens overlay\n            \n", "")]
  aut[, name := str_replace(name, "\n", "")]
  aut[, email := str_replace(email, "mailto:", "")]
  aut[, rank := .I]
  return(aut)
}

```

```{r, eval=F}

# Build a complete list of authors and affiliations
authors <- scrapeAuthor(refs[1, url], refs[1, uid])
bad <- integer(0)

for (i in 2:nrow(refs)) {
  cat(i, "\n")
  tmp <- try(scrapeAuthor(refs[i, url], refs[i, uid]))
  if( class(tmp)=="try-error") {
    # Beep and mark bad records
    alarm() 
    bad <- c(bad, i)
  } else authors <- rbind(authors, tmp)
}

# Try the bad ones again
bad2 <- integer(0)
for (i in bad) {
  cat(i, "\n")
  tmp <- try(scrapeAuthor(refs[i, url], refs[i, uid]))
  if( class(tmp)=="try-error") {
    alarm() 
    bad2 <- c(bad2, i)
  } else authors <- rbind(authors, tmp)
}

# Save and mark records that could not be scraped
bad <- bad2; rm(bad2)
refs[, hasAuthor := T]
refs[bad, hasAuthor := F]

# We seem to be missing names of authors for 10 publications
bad_name <- authors[is.na(name), unique(uid)]

# Let's try to scrape these pages again
authors <- authors[!uid %in% bad_name]
bad2 <- integer(0)
for (i in bad_name) {
  cat(i, "\n")
  tmp <- try(scrapeAuthor(refs[uid==i, url], i))
  if( class(tmp)=="try-error") {
    alarm() 
    bad2 <- c(bad2, i)
  } else authors <- rbind(authors, tmp)
}

# => ignore the 4 pubs that cannot be scraped for author names
authors <- authors[!is.na(name)]
refs[uid %in% bad_name, hasAuthor := F]

# Consolidate the list of authors, make unique author identifiers.
# Note that some authors have an `orcid` of the form `authname_N8eaef220N8ea76980`
# but that's not in a typical ORCID format (should be 16-digit), must be Scopus author ID or
# hash tag; not sure it's usable at all.

# Make arbitrary unique author id for now
authors[url %like% "author-", url := NA]
authors[, name := str_trim(name)]
authors[, email := str_trim(email)]
authors[, aff := str_trim(aff)]
authors[, authid := paste0("aut-", str_pad(.GRP, 5, pad="0")), keyby=name]

# Remove duplicated author by publication uid
setkey(authors, uid, authid)
authors <- unique(authors)

# Some affiliations are clearly bad
authors[aff %like% "Chapter", aff := NA]

setcolorder(authors, c(8,1,2:7))
setcolorder(refs, c(23, 25, 1:22, 27:29, 26, 24))

```

```{r}

# Print authors with same name but different affiliation
tmp <- authors[, .(name=unique(name)), keyby=.(authid, aff)]
tmp[, N:=.N, by=name]
pander(tmp[N>1][order(name)][1:21])

```

The steps above return:

* `r nrow(authors)` authors/co-authors across all
CGIAR publications.
* At first glance there are `r authors[, uniqueN(authid)]` unique researchers (authors with unique names and ScopusID), but this might need further cleaning. 
* `r authors[!is.na(url), .N]` researchers have an
ORCID/ScopusID, and `r authors[!is.na(email), .N]` authors show an e-mail
address. Note that author ScopusID, when present,
could be used to augment information about known authors (e.g. impact factor,
location, etc.).
* Out of the `r authors[, uniqueN(authid)]` unique researchers, we have affiliation information for `r authors[!is.na(aff), uniqueN(authid)]`.
* Out of the `r nrow(refs)` CGIAR publication records from
ScienceDirect author details could not be extracted for `r length(bad)+length(bad_name)` entries (server error or ill-formatted pages). 

# Improve author affiliation

Before looking at trends in co-authorships we need to identify
authors affiliated with CGIAR centers. In principle each publication should have
at least 1 affiliated author. We could also try to disambiguate affiliation names, and then geolocate them.


```{r, eval=F}

# Try to identify CGIAR-affiliated authors
for (i in unique(cg)) authors[aff %like% i, institution := i]

# And a few more manual recodes
authors[aff %like% "Bioversity", institution := "Bioversity International"]
authors[aff %like% "INIBAP", institution := "Bioversity International"]
authors[aff %like% "ILRAD", institution := "ILRI"]
authors[aff %like% "ILCA", institution := "ILRI"]
authors[aff %like% "International Livestock Research Institute", institution := "ILRI"]
authors[aff %like% "World Agroforestry", institution := "ICRAF"]
authors[aff %like% "International Potato Center", institution := "CIP"]
authors[aff %like% "International Rice Research Institute", institution := "IRRI"]
authors[aff %like% "ISNAR", institution := "IFPRI"]
authors[aff %like% "International Food Policy Research Institute", institution := "IFPRI"]
authors[aff %like% "International Maize and Wheat", institution := "CIMMYT"]
authors[aff %like% "International Institute for Tropical Agriculture", institution := "IITA"]
authors[aff %like% "International Institute of Tropical Agriculture", institution := "IITA"]
authors[aff %like% "Centro Internacional de Agricultura Tropical", institution := "CIAT"]
authors[aff %like% "International Center for Tropical Agriculture", institution := "CIAT"]
authors[tolower(aff) %like% "centre for agricultural research in the dry areas", institution := "ICARDA"]
authors[tolower(aff) %like% "institute for the semi-arid tropics", institution := "ICRISAT"]
authors[tolower(aff) %like% "international forestry research", institution := "CIFOR"]
authors[tolower(aff) %like% "international water management institute", institution := "IWMI"]

# Are there publications where no author is marked as CGIAR?
tmp <- authors[, .(N=sum(!is.na(institution), na.rm=T)), by=uid][N==0]
# => still 1073 entries
tmp <- authors[uid %in% tmp$uid, .(authid, name, aff), keyby=uid]
# => they seem to correspond to manuscripts with large list of authors, not clear why they 
# appear in ScienDirect query results.

# Indicate which publications have known CGIAR-affiliated authors
tmp <- authors[, .(N=sum(!is.na(institution), na.rm=T)), by=uid]
setkey(refs, uid)
setkey(tmp, uid)
refs[tmp, nCGIARAuthor := N]

# Export metadata
refs.lbl <- c(
  "unique publication ID (arbitrary)",
  "unique publication ID (ScienceDirect)",
  "publication type",
  "title",
  "journal",
  "volume",
  "number",
  "pages",
  "publication year",
  "ISSN",
  "DOI",
  "ScienceDirect URL",
  "concat list of authors",
  "keywords",
  "abstract",
  "editor",
  "book title",
  "publisher",
  "series",
  "edition",
  "publisher city or country",
  "ISBN",
  "notes",
  "CGIAR center affiliation",
  "affiliation of 1st author",
  "affiliation country of 1st author",
  "affiliation city of 1st author",
  "number of citations",
  "has matching authors (TRUE/FALSE)",
  "count of matching CGIAR-affiliated authors")

authors.lbl <- c(
  "unique author ID (arbitrary)",
  "unique publication ID (matching list of publications)",
  "full name",
  "ScienceDirect URL",
  "e-mail",
  "affiliation ID on web page (ignore)",
  "affiliation",
  "author rank in publication",
  "CGIAR center affiliation")

refs.lbl <- data.table(varCode=names(refs), varLabel=refs.lbl)
authors.lbl <- data.table(varCode=names(authors), varLabel=authors.lbl)

# Export to CSV
write.csv(refs, "./out/2016.10/sciencedirect_pubs.csv", na="", row.names=F)
write.csv(authors, "./out/2016.10/sciencedirect_authors.csv", na="", row.names=F)
write.csv(refs.lbl, "./out/2016.10/sciencedirect_pubs_meta.csv", na="", row.names=F)
write.csv(authors.lbl, "./out/2016.10/sciencedirect_authors_meta.csv", na="", row.names=F)

```


```{r}

# Tally identified CGIAR authors by publication
pander(table(refs$nCGIARAuthor))


# Tally the number of pairwise authorships to draw network graph


# Add count of CGIAR-affiliated publications to each author

# Add list of unique keywords to each author

# Tally unique authors by CGIAR-affiliation


```

# Graph authorships


```{r, eval=F}

# Prepare author data for graphing
nodes <- authors
nodes[, ID := 0:(.N-1)]
nodes[, size := scale(cited, center=F)*4]
nodes[, group := factor(cgiar, labels=c("Partners", "CGIAR"))]
links <- cg.coauthor[id %in% nodes$id]
links <- links[ids %in% nodes$id]
setkey(links, id)
links$from <- nodes[links][, ID]
setkey(links, ids)
links$to <- nodes[links][, ID]
links[, value := 3]
setcolorder(nodes, c(8:10, 1:7))
setcolorder(links, c(4,5,6,1:3))

# Prepare CGIAR center data for graphing


```


# Next steps

* Find author impact factors and geographic location using ORCID or ScopusID?
* Possible semantic analysis on article titles, abstracts, and keywords
  using **MonkeyLearn API** (see https://github.com/ropenscilabs/monkeylearn
  and http://monkeylearn.com/) to identify key concepts and terms (and help
  build ANT).
* Identify place names from titles, abstracts, and keywords (CIAT)


```{r save, eval=F}

# Save snapshot
rm(list=ls()[ls() %like% "tmp"])
rm(x, y, i, j, aff, aut, q)
save.image("./tmp/ScienceDirect.RData")

```

